---
title: "p8105_hw2_yj2931"
author: "Yuyao"
date: "2026-10-06"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Problem one

Steps: 
1. Read and clean the data
2. Retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. 
3. Convert the entry variable from character (YES vs NO) to a logical variable

```{r}
nyc_transit_df = 
  read.csv(file = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("","NA", ".")) |> 
  janitor::clean_names() |> 
  
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
 
   mutate(
    entry = case_match(
    entry,
    "YES" ~ TRUE,
    "NO" ~ FALSE
  )
)
```
### Description:

The dataset contains information on NYC subway entrances and exits, including line, station name, station latitude and longitude, routes served, entry status, vending availability, entrance type, and ADA compliance. 

I cleaned the variable names, kept only the variables required for this problem, and converted `entry` from `"YES"` / `"NO"` character values to a logical variable. 

The resulting dataset has `r nrow(nyc_transit_df)` rows and `r ncol(nyc_transit_df)` columns. The data are not fully tidy because the routes served are stored across multiple columns (`route1` to `route11`) instead of in one route variable.

### 1. Number of distinct stations
```{r}
num_distinct = nyc_transit_df |>
  distinct(line, station_name) |>
  nrow()
```

### 2. Number of ADA-compliant stations

```{r}
num_ada = nyc_transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

### 3. Proportion of entrances/exits without vending that allow entrance

```{r}
proportion_no_vending = nyc_transit_df |>
  filter(vending == "NO") |>
  summarize(
    proportion = round(mean(entry, na.rm = TRUE), 2)
  )
```

There are `r num_distinct` stations here. `r num_ada` stations are ADA compliant. The proportion of station entrances / exits without vending allow entrance is around `r proportion_no_vending` (rounded to 2 decimal places).
