---
title: "p8105_hw2_yj2931"
author: "Yuyao"
date: "2026-10-06"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

## Problem one

Steps: 
1. Read and clean the data
2. Retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. 
3. Convert the entry variable from character (YES vs NO) to a logical variable

```{r}
nyc_transit_df = 
  read.csv(file = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("","NA", ".")) |> 
  janitor::clean_names() |> 
  
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
 
   mutate(
    entry = case_match(
    entry,
    "YES" ~ TRUE,
    "NO" ~ FALSE
  )
)
```
### Description:

The dataset contains information on NYC subway entrances and exits, including line, station name, station latitude and longitude, routes served, entry status, vending availability, entrance type, and ADA compliance. 

I cleaned the variable names, kept only the variables required for this problem, and converted `entry` from `"YES"` / `"NO"` character values to a logical variable. 

The resulting dataset has `r nrow(nyc_transit_df)` rows and `r ncol(nyc_transit_df)` columns. The data are not fully tidy because the routes served are stored across multiple columns (`route1` to `route11`) instead of in one route variable.

### 1. Number of distinct stations
```{r}
num_distinct = count(distinct(nyc_transit_df, station_name, line))
```

### 2. Number of ADA-compliant stations

```{r}
num_ada = nyc_transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

### 3. Proportion of entrances/exits without vending that allow entrance

```{r}
proportion_no_vending = nyc_transit_df |>
  filter(vending == "NO") |>
  summarize(
    proportion = round(mean(entry, na.rm = TRUE), 2)
  )
```

Reformat data so that route number and route name are distinct variables. 

```{r}
nyc_transit_df =
  nyc_transit_df |>
  mutate(
    across(route1:route11, as.character)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route"
  ) |>
  drop_na(route_name)
```


There are `r num_distinct` stations here. `r num_ada` stations are ADA compliant. The proportion of station entrances / exits without vending allow entrance is around `r proportion_no_vending` (rounded to 2 decimal places).


## Problem 2

Import and clean the dataset.

```{r}
mr_trash_wheel_df = 
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743",
    skip = 1
  ) |>
  janitor::clean_names()
```

Omit rows that do not include dumpster-specific data.
Round the number of sports balls to the nearest integer and converts the result to an integer variable.

```{r}
mr_trash_wheel_df = 
  mr_trash_wheel_df |> 
    filter(!is.na(dumpster)) |> 
    mutate(
      sports_balls = as.integer(round(sports_balls)),
      trash_wheel = "Mr. Trash Wheel",
      year = as.integer(year),
      plastic_bottles = parse_number(as.character(plastic_bottles))
    )
```

Import the data for Professor Trash Wheel. 

```{r}
professor_trash_wheel_df = 
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    skip = 1
  ) |>
  janitor::clean_names() |> 
      filter(!is.na(dumpster),
             plastic_bottles != "dive") |> 
  mutate(
    trash_wheel = "Professor Trash Wheel",
    year = as.integer(year),
    plastic_bottles = parse_number(as.character(plastic_bottles))
  )
```

Import the data for Gwynnda. 

```{r}
gwynnda_df = 
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395",
    skip = 1
  ) |>
  janitor::clean_names() |> 
      filter(!is.na(dumpster)) |> 
  mutate(
    trash_wheel = "Gwynnda",
    year = as.integer(year),
    plastic_bottles = parse_number(as.character(plastic_bottles))
  )
```

Combine with the Mr. Trash Wheel dataset to produce a single tidy dataset.

```{r}
trash_wheel_tidy_df =
  bind_rows(
    mr_trash_wheel_df,
    professor_trash_wheel_df,
    gwynnda_df
  )
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.

For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?





